---
title: "ChrX global biallelic status in Female and Male - Deng AR"
author: "Paolo Martini"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We load the data we downloaded from the processed set in GSE45719. Some labels are duplicated and the second occurrence has been removed.

```{r}
data <- read.table("Deng-alleic-expression-for-X-genes.txt.gz", header=T, sep="\t", 
                   check.names = F, stringsAsFactors = F)

duplicated <- unique(data$label[which(duplicated(data$label))])
data <- data[!(data$label %in% duplicated), , drop=F]
row.names(data) <- data$label
data$label <- NULL

valid_genes <- read.table("genes_shared_in_chrX.txt", sep="\t", header=F, stringsAsFactors = F)$V1
data <- data[valid_genes, , drop=F]
if (any(is.na(row.names(data)))) stop()
```

From column names we are going to retrieve the stage.

```{r}
type <- sapply(strsplit(colnames(data), "_"), function(x) x[1])
sorted_stages <- c("zy1", "zy2", "zy3", "zy4", "early2cell", "late2cell", "mid2cell", "4cell", "8cell", "16cell",  "earlyblast", "midblast", "lateblast", "fibroblast", "BXC", "C57twocell")
setdiff(unique(type), sorted_stages)

type <- factor(type, levels=sorted_stages)
column_order <- order(type)
pheno <- data.frame(cns = colnames(data), type = type, stringsAsFactors = F)[column_order, ]

```

We retrieve by email of the author the sex of some samples.

```{r}
samples <- read.table(file="sample-annotation.txt", header=F, sep="\t", stringsAsFactors = F, check.names = F)

female_samples <- samples$V1[samples$V2=="female"]
samples_female <- unique(unlist(lapply(female_samples, function(sample) {
  idx <- grep(sample, as.character(pheno$cns))
  as.character(pheno$cns)[idx]
})))

samples_female_fct <- do.call(rbind, lapply(female_samples, function(sample) {
  idx <- grep(sample, as.character(pheno$cns))
  pheno[idx, ]
}))
levels(samples_female_fct$type)

male_samples <- samples$V1[samples$V2=="male"]
samples_male <- unique(unlist(lapply(male_samples, function(sample) {
  idx <- grep(sample, as.character(pheno$cns))
  as.character(pheno$cns)[idx]
})))

samples_male_fct <- do.call(rbind, lapply(male_samples, function(sample) {
  idx <- grep(sample, as.character(pheno$cns))
  pheno[idx, ]
}))

drop = (grepl("_smartseq2", as.character(samples_male_fct$cns)) | grepl("pooled_split", as.character(samples_male_fct$cns)))

samples_male_fct = samples_male_fct[!drop, ]
samples_male_fct_dev_female <- samples_male_fct[samples_male_fct$type %in% samples_female_fct$type, ]

all_samples <- c(samples_male_fct_dev_female$cns, samples_female_fct$cns)
```

Now we shrink the data accordingly.

```{r}
data <- data[, all_samples, drop=F]
```

We can compute the ratio _(ratio of the first in the total)_

```{r}
ratios_global_tot <- sapply(data, function(x) {
  mat <- do.call(rbind, strsplit(x, ","))
  sum(as.numeric(mat[,2]))/sum(as.numeric(mat[,1]))
})
```

```{r}
library(ggplot2)

female <- data.frame(meanAR = ratios_global_tot[samples_female_fct$cns],
                     group="Female", stage=samples_female_fct$type)
female <- droplevels(female)
female <- tapply(female$meanAR, female$stage, mean)
female <- data.frame(meanAR=female, stage=names(female), group="Female")

male <- data.frame(meanAR = ratios_global_tot[samples_male_fct_dev_female$cns], group="Male", stage=samples_male_fct_dev_female$type)
male <- droplevels(male)
male <- tapply(male$meanAR, male$stage, mean)
male <- data.frame(meanAR=male, stage=names(male), group="Male")

overall_mean_AR <- rbind(female, male)
overall_mean_AR$stage <- factor(as.character(overall_mean_AR$stage), levels=c("mid2cell", "late2cell", "4cell", "16cell", "earlyblast"))

ggplot(overall_mean_AR, aes(x=stage, y=meanAR, group=group, color=group)) +
  geom_line() +
  theme_classic() +
  ylim(c(0,1))

```

# Create heatmap of AR

```{r}
ratios <- sapply(data, function(x) {
  sapply(strsplit(x, ","), function(y) {
    ac <- as.numeric(y)
    if (all(ac==0))
      return(NA)
    ac[1]/sum(ac)
  })
})
row.names(ratios) <- row.names(data)
```

We remove the line with all NAs.

```{r}
ratios <- ratios[!apply(is.na(ratios), 1, all), ]
```

```{r}
gene_positions_ad <- read.table("symbol2tss.txt", header=F, check.names = F, stringsAsFactors = F, strip.white = F)
colnames(gene_positions_ad) <- c("symbol", "chr", "tss")

female_data <- ratios[, samples_female_fct$cns]
female_data <- data.frame(gene=row.names(female_data), female_data, check.names = F)

xgene_positions <- gene_positions_ad[gene_positions_ad$symbol %in% row.names(female_data),  ]
sort_x_genes <- xgene_positions$symbol[order(xgene_positions$tss)]

female_data$gene <- factor(as.character(female_data$gene), levels = sort_x_genes)

dataset <- tidyr::gather(female_data, cell_stage, AR, -c(gene))
dataset$cell_stage <- factor(dataset$cell_stage, levels=rev(samples_female_fct$cns))

colors <- RColorBrewer::brewer.pal(9, "Set1")
low = colors[1]
mid = colors[6]
high = colors[1]

labels <- rev(as.character(samples_female_fct$type))
labels[duplicated(labels)] <- ""
ticks_colors <- rep("black", length(labels))
ticks_colors[labels==""] <- "transparent"

labels <- c("", labels[-c(length(labels))])

s_fig_6a <- ggplot(dataset, aes(y=cell_stage, x=gene, color=AR)) +
  theme_classic() +
  geom_point(size=ifelse(is.na(dataset$AR), 0.1, 1)) +
  theme(axis.line=element_blank(), 
        axis.text.y = element_text(size=6),
        axis.text.x = element_text(size=2, angle=90), axis.ticks = element_blank(),
        axis.ticks.x = element_blank()) +
  scale_colour_gradient2(low = low, mid = mid, high = high, midpoint = 0.5, space = "Lab",
                         na.value = colors[9], guide = "colourbar", aesthetics = "color") +
  labs(color = "AR", y="Stages of mouse development", x="Chromosome X genes") +
  scale_y_discrete(labels=labels) +
  theme(axis.ticks.y = element_line(color=ticks_colors))

s_fig_6a
```